---
title: "Week 3 Homework"
author: "Aaron"
date: "June 2, 2016"
output:
  html_document:
    highlight: haddock
    theme: cosmo
  pdf_document: default
---

### Part 1:  Rat Weight Gain

#### Read Data and Process Accordingly

```{r message=FALSE}

library(dplyr)
library(readr)
library(knitr)
library(magrittr)

```


```{r}

rat <- read_csv("../../../../../Downloads/RatWeightGain.csv")

```

Check what the data looks like:

```{r}
glimpse(rat)
```

#### Calculate means & standard deviations for 4 groups broken down by source and type

The mean and standard deviation are listed in the table below:

```{r}

rat %>% 
        group_by(SOURCE,TYPE) %>% 
        summarise(mean = mean(WEIGHTGAIN), 
                  std = sd(WEIGHTGAIN)) %>% kable()

```

#### Perform oneway tests twice: once for source and once for type

1. test for `SOURCE`:

```{r}

rat %>% 
        oneway.test(WEIGHTGAIN ~ SOURCE, ., var.equal=T)

```

The mean of Weight Gain between source groups are **not** significantly different. 

2. test for `TYPE`:

```{r}

rat %>% 
        oneway.test(WEIGHTGAIN ~ TYPE, ., var.equal=T)

```

The mean of Weight Gain between source groups are significantly different. 

#### Perform two way test

```{r}

rat %>% 
        lm(WEIGHTGAIN ~ TYPE + SOURCE, .) %>% 
        anova()

```

There is likely a significant interaction between two variables. 

### Part 2:  mtcars

Use the `table()` function with the following combinations

- The variables vs and am
- The variables gear and carb
- The variables cyl and gear

For each of the three cases above guess what the results of a Chi-Squared analysis will be 


Perform a Chi-Squared analysis on the mtcars data set for each of the three cases above

```{r warning=F}
# vs & am

mtcars %$% 
        table(vs, am) %>% 
        chisq.test()
```

The counts are likely due to chance. 

```{r warning=F}
# gear & carb

mtcars %$% 
        table(gear, carb) %>% 
        chisq.test()
```

The counts are likely due to chance, but less likely than previous pair. 

```{r warning=F}
# cyl & gear

mtcars %$% 
        table(cyl, gear) %>% 
        chisq.test()
```

The counts are **not** likely due to chance. 
